---
output:
  word_document: default
  html_document: default
---


## S19 BAN502 Course Project Part 1 
## Data Exploration, Preparation and Visualization
### Michael Duncan


libraries
```{r}
options(tidyverse.quiet = TRUE)
library(tidyverse)
library(psych) # For Descriptive Statistics
library(mice) #package for imputation
library(VIM) #visualizing missingness
library(ggcorrplot)
library(GGally)
```
Read Rain Data Set
```{r}
rain <- read_csv("rain.csv")
glimpse(rain)

```

Convert to factors
```{r}
#updated all character variables to factors
#note when imputating discoverd that R does not like character values at all. 
rain2 <- rain %>%
  mutate(RainToday = as_factor(as.character(RainToday))) %>%
  mutate(RainTomorrow = as_factor(as.character(RainTomorrow))) %>%
  mutate(WindGustDir = as_factor(as.character(WindGustDir))) %>%
  mutate(WindDir9am = as_factor(as.character(WindDir9am))) %>%
  mutate(WindDir3pm = as_factor(as.character(WindDir3pm)))
glimpse(rain2)
```

Summary
```{r}
summary(rain2)
```

View missingness  
```{r}
vim_plot = aggr(rain2, numbers = TRUE, prop = c(TRUE, FALSE),cex.axis=.7)
#the cex.axis reduces size of text on x-axis so labels fit better
```

imputation  
```{r}
#select only variables relevant to our analysis
#Chose all variables because they could all have some level of predictive power
rain2 = rain2 %>% select(c("Date","MinTemp", "MaxTemp", "Rainfall", "WindGustDir", "WindGustSpeed", "WindDir9am", "WindDir3pm", "WindSpeed9am", "WindSpeed3pm", "Humidity9am", "Humidity3pm","Pressure9am", "Pressure3pm", "Cloud9am", "Cloud3pm", "Temp9am", "Temp3pm", "RainToday", "RainTomorrow"))

imp_testtest = mice(rain2, m=1, method='pmm', printFlag=FALSE)
summary(imp_testtest)
```


```{r}
densityplot(imp_testtest, ~MinTemp) #red imputed, blue original
densityplot(imp_testtest, ~MaxTemp) #red imputed, blue original
densityplot(imp_testtest, ~Rainfall) #red imputed, blue original
densityplot(imp_testtest, ~WindGustDir) #red imputed, blue original
densityplot(imp_testtest, ~WindGustSpeed) #red imputed, blue original
densityplot(imp_testtest, ~WindDir9am) #red imputed, blue original
densityplot(imp_testtest, ~WindDir3pm) #red imputed, blue original
densityplot(imp_testtest, ~WindSpeed9am) #red imputed, blue original
densityplot(imp_testtest, ~WindSpeed3pm) #red imputed, blue original
densityplot(imp_testtest, ~Humidity9am) #red imputed, blue original
densityplot(imp_testtest, ~Humidity3pm) #red imputed, blue original
densityplot(imp_testtest, ~Pressure9am) #red imputed, blue original
densityplot(imp_testtest, ~Pressure3pm) #red imputed, blue original
densityplot(imp_testtest, ~Cloud9am) #red imputed, blue original
densityplot(imp_testtest, ~Cloud3pm) #red imputed, blue original
densityplot(imp_testtest, ~Temp9am) #red imputed, blue original
densityplot(imp_testtest, ~Temp3pm) #red imputed, blue original
densityplot(imp_testtest, ~RainToday) #red imputed, blue original





```
Merge the imputed values into our rain2 data frame  
```{r}
rain3 <- complete(imp_testtest) 
summary(rain3)
glimpse(rain3)
```

View missingness  of updated data frame rain3
```{r}
vim_plot = aggr(rain3, numbers = TRUE, prop = c(TRUE, FALSE),cex.axis=.7)
#the cex.axis reduces size of text on x-axis so labels fit better
```



###Looking for Outliers

Visualization Quantitative Variables --- Histograms
```{r}
ggplot(rain3, aes(x=MinTemp )) + geom_histogram() + theme_bw()
describe(rain$MinTemp) # From Psych package to give summary statistics
```

```{r}
ggplot(rain3, aes(x=MaxTemp )) + geom_histogram() + theme_bw()
describe(rain$MaxTemp) # From Psych package to give summary statistics
```


```{r}
ggplot(rain3, aes(x=Rainfall )) + geom_histogram() + theme_bw()
describe(rain$Rainfall)
```

```{r}
#rain4 <- rain3 %>% filter(Rainfall < 36) ## Testing this removes aroun 1.1% of the data This is 3 SDev of the mean
#ggplot(rain4, aes(x=Rainfall)) + geom_histogram()
#summary(rain4)
#sd(rain3$Rainfall)
#describe(rain$Rainfall) # From Psych package to give summary statistics
```





```{r}
ggplot(rain3, aes(x=WindGustSpeed )) + geom_histogram() + theme_bw()
describe(rain3$WindGustSpeed)
```

```{r}
ggplot(rain3, aes(x=WindSpeed9am )) + geom_histogram() + theme_bw()
describe(rain3$WindSpeed9am)
```

```{r}
ggplot(rain3, aes(x=WindSpeed3pm )) + geom_histogram() + theme_bw()
describe(rain3$WindSpeed3pm)
```

```{r}
ggplot(rain3, aes(x=Humidity9am )) + geom_histogram() + theme_bw()

```

```{r}
ggplot(rain3, aes(x=Humidity3pm )) + geom_histogram() + theme_bw()

```

```{r}
ggplot(rain3, aes(x=Pressure9am )) + geom_histogram() + theme_bw()
describe(rain3$Pressure9am)
```

```{r}
ggplot(rain3, aes(x=Pressure3pm )) + geom_histogram() + theme_bw()
describe(rain3$Pressure3pm)
```

```{r}
ggplot(rain3, aes(x=Cloud9am )) + geom_histogram() + theme_bw()

```

```{r}
ggplot(rain3, aes(x=Cloud3pm )) + geom_histogram() + theme_bw()

```

```{r}
ggplot(rain3, aes(x=Temp9am )) + geom_histogram() + theme_bw()
describe(rain3$Temp9am)
```

```{r}
ggplot(rain3, aes(x=Temp3pm )) + geom_histogram() + theme_bw()
```

### Visualizations

Visualizations Quantatitive Variables --- Box Plots
```{r}
ggplot(rain3, aes(x=RainTomorrow, y= MinTemp)) + geom_boxplot() + theme_bw()
ggplot(rain3, aes(x=RainTomorrow, y= MaxTemp)) + geom_boxplot() + theme_bw()
ggplot(rain3, aes(x=RainTomorrow, y= Rainfall)) + geom_boxplot() + theme_bw()
ggplot(rain3, aes(x=RainTomorrow, y= WindGustSpeed)) + geom_boxplot() + theme_bw()
ggplot(rain3, aes(x=RainTomorrow, y= WindSpeed9am)) + geom_boxplot() + theme_bw()
ggplot(rain3, aes(x=RainTomorrow, y= WindSpeed3pm)) + geom_boxplot() + theme_bw()
ggplot(rain3, aes(x=RainTomorrow, y= Humidity9am)) + geom_boxplot() + theme_bw()
ggplot(rain3, aes(x=RainTomorrow, y= Humidity3pm)) + geom_boxplot() + theme_bw()
ggplot(rain3, aes(x=RainTomorrow, y= Pressure9am)) + geom_boxplot() + theme_bw()
ggplot(rain3, aes(x=RainTomorrow, y= Pressure3pm)) + geom_boxplot() + theme_bw()
ggplot(rain3, aes(x=RainTomorrow, y= Cloud9am)) + geom_boxplot() + theme_bw()
ggplot(rain3, aes(x=RainTomorrow, y= Cloud3pm)) + geom_boxplot() + theme_bw()
ggplot(rain3, aes(x=RainTomorrow, y= Temp9am)) + geom_boxplot() + theme_bw()
ggplot(rain3, aes(x=RainTomorrow, y= Temp3pm)) + geom_boxplot() + theme_bw()


```

```{r}
ggplot(rain3, aes(x=RainTomorrow, y= Cloud3pm)) + geom_boxplot() + theme_bw()
ggsave("Cloud3pm.png")
```




Visualizations of Categorical Variables
```{r}
ggplot(rain3, aes(x=RainToday, fill = RainTomorrow)) + geom_bar(position ="fill") + theme_bw()
ggplot(rain3, aes(x=WindGustDir, fill = RainTomorrow)) + geom_bar(position ="fill") + theme_bw()
ggplot(rain3, aes(x=WindDir9am, fill = RainTomorrow)) + geom_bar(position ="fill") + theme_bw()
ggplot(rain3, aes(x=WindDir3pm, fill = RainTomorrow)) + geom_bar(position ="fill") + theme_bw()
```





